---
title: "Space-Time Prediction of Bike Share Demand: Philadelphia Indego"
author: "Xiao Yu, Qianmu Zheng"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

## Assignment Structure

- Download a different quarter of Indego data (Q4 2024)
- Apply the same workflow
- Analyze error patterns in your quarter
- Add 2-3 new features to improve the model
- Write a brief report on what you learned

---

# Setup

## Load Libraries

```{r load_libraries}
# Core tidyverse
library(tidyverse)
library(lubridate)

# Spatial data
library(sf)
library(tigris)

# Census data
library(tidycensus)

# Weather data
library(riem)  # For Philadelphia weather from ASOS stations

# Visualization
library(viridis)
library(gridExtra)
library(knitr)
library(kableExtra)

# here!
library(here)
# Get rid of scientific notation. We gotta look good!
options(scipen = 999)
```

## Define Themes

```{r themes}
plotTheme <- theme(
  plot.title = element_text(size = 14, face = "bold"),
  plot.subtitle = element_text(size = 10),
  plot.caption = element_text(size = 8),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title = element_text(size = 11, face = "bold"),
  panel.background = element_blank(),
  panel.grid.major = element_line(colour = "#D0D0D0", size = 0.2),
  panel.grid.minor = element_blank(),
  axis.ticks = element_blank(),
  legend.position = "right"
)

mapTheme <- theme(
  plot.title = element_text(size = 14, face = "bold"),
  plot.subtitle = element_text(size = 10),
  plot.caption = element_text(size = 8),
  axis.line = element_blank(),
  axis.text = element_blank(),
  axis.ticks = element_blank(),
  axis.title = element_blank(),
  panel.background = element_blank(),
  panel.border = element_blank(),
  panel.grid.major = element_line(colour = 'transparent'),
  panel.grid.minor = element_blank(),
  legend.position = "right",
  plot.margin = margin(1, 1, 1, 1, 'cm'),
  legend.key.height = unit(1, "cm"),
  legend.key.width = unit(0.2, "cm")
)

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

## Set Census API Key

```{r census_key, eval=FALSE}

census_api_key("ec702835845a134b4376c60759aa72ce62f6df59", overwrite = TRUE, install = TRUE)

```

```{r census_key_hidden, include=FALSE}
# Hidden key for rendering
census_api_key("ec702835845a134b4376c60759aa72ce62f6df59", overwrite = TRUE)
```

---

# Data Import & Preparation

## Load Indego Trip Data (Q4 2024)

```{r load_indego}
indego <- read_csv("/Users/cathy/stats a1/Untitled/musa-5080-fall-2025-portfolio-setup-student-portfolio-kc2/Assignments/Assignment_last/indego-trips-2024-q4.csv")

getwd()


# Quick look at the data
glimpse(indego)
```

## Examine the Data Structure

```{r explore_data}
# How many trips?
cat("Total trips in Q4 2024:", nrow(indego), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indego$start_time)), "to", 
    max(mdy_hm(indego$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indego$start_station)), "\n")

# Trip types
table(indego$trip_route_category)

# Passholder types
table(indego$passholder_type)

# Bike types
table(indego$bike_type)
```

## Create Time Bins

We need to aggregate trips into hourly intervals for our panel data structure.

```{r create_time_bins}
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

---

# Exploratory Analysis

## Trips Over Time

```{r trips_over_time}
# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q4 2024",
    subtitle = "Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```


## Hourly Patterns

```{r hourly_patterns}
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

## Top Stations

```{r top_stations}
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Get Philadelphia Spatial Context

## Load Philadelphia Census Data

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  # WGS84 for lat/lon matching

# Check the data
glimpse(philly_census)
```

## Map Philadelphia Context

```{r map_philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

## Join Census Data to Stations

```{r join_census_to_stations}
# Create sf object for stations
stations_sf <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme



```

# Dealing with missing data

We need to decide what to do with the non-residential bike share stations. For this example, we are going to remove them -- this is not necessarily the right way to do things always, but for the sake of simplicity, we are narrowing our scope to only stations in residential neighborhoods. We might opt to create a separate model for non-residential stations..


```{r}
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```


# Get Weather Data

```{r get_weather}
# Get weather from Philadelphia International Airport (KPHL)
weather_data <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2024-10-01",
  date_end = "2024-12-31"
)

# Process weather data
weather_processed <- weather_data %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct()

# Check for missing hours and interpolate if needed
weather_complete <- weather_processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete %>% select(Temperature, Precipitation, Wind_Speed))
```

## Visualize Weather Patterns

```{r visualize_weather}
ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q4 2024",
    subtitle = "Winter to early spring transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

---

# Create Space-Time Panel

## Aggregate Trips to Station-Hour Level

```{r aggregate_trips}
# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))
```

## Create Complete Panel Structure

Not every station has trips every hour. We need a **complete panel** where every station-hour combination exists (even if Trip_Count = 0).

```{r complete_panel}
# Calculate expected panel size
n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand.grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel, by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

# Fill in station attributes (they're the same for all hours)
station_attributes <- trips_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

## Add Time Features

```{r add_time_features}
study_panel <- study_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE, locale = "C"),
    dotw = wday(interval60, label = TRUE, locale = "C"),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

```

## Join Weather Data

```{r join_weather}
study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60")

# Check for missing values
summary(study_panel %>% select(Trip_Count, Temperature, Precipitation))
```

---

# Create Temporal Lag Variables

The key innovation for space-time prediction: **past demand predicts future demand**.

## Why Lags?

If there were 15 bike trips from Station A at 8:00 AM, there will probably be ~15 trips at 9:00 AM. We can use this temporal persistence to improve predictions.

```{r create_lags}
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

## Visualize Lag Correlations

```{r lag_correlations}
# Sample one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

---

# Temporal Train/Test Split

**CRITICAL:** We must train on PAST data and test on FUTURE data!

## Why Temporal Validation Matters

In real operations, at 6:00 AM on March 15, we need to predict demand for March 15-31. We have data from Jan 1 - March 14, but NOT from March 15-31 (it hasn't happened yet!).

**Wrong approach:** Train on weeks 10-13, test on weeks 1-9 (predicting past from future!)

**Correct approach:** Train on weeks 1-9, test on weeks 10-13 (predicting future from past)

```{r}
# Split by week

# Which stations have trips in BOTH early and late periods?
early_stations <- study_panel_complete %>%
  filter(week < 48) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(week >= 48) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations <- intersect(early_stations, late_stations)


# Filter panel to only common stations
study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

study_panel_complete <- study_panel_complete %>%
  mutate(
    month = lubridate::month(date, label = TRUE),
    month = factor(month),
    start_station = factor(start_station),
    dotw_simple = factor(
      dotw,
      levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
    )
  )
```

```{r}
library(slider)

study_panel_complete <- study_panel_complete %>%
  arrange(start_station, date, hour) %>%
  group_by(start_station) %>%

  mutate(
    # Perfect biking weather
    perfect_weather = ifelse(Temperature >= 60 & Temperature <= 75 & Precipitation == 0, 1, 0),

    # Rolling 7-day average demand (168 hours)
    rolling7 = slide_dbl(Trip_Count, mean, .before = 24*7, .complete = FALSE),

    # Same hour last week
    same_hour_last_week = lag(Trip_Count, 24*7)
  ) %>%
  ungroup()

study_panel_complete <- study_panel_complete %>%
  mutate(
    rolling7 = ifelse(is.na(rolling7), mean(Trip_Count, na.rm = TRUE), rolling7),
    same_hour_last_week = ifelse(is.na(same_hour_last_week),
                                 mean(Trip_Count, na.rm = TRUE),
                                 same_hour_last_week)
  )

```


```{r}
train <- study_panel_complete %>%
  filter(week < 48)

test <- study_panel_complete %>%
  filter(week >= 48)

cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")


cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")

```

```{r}
train <- train %>%
  mutate(
    month         = factor(month),
    start_station = factor(start_station),
    dotw_simple   = factor(dotw_simple)
  )

test <- test %>%
  mutate(
    month         = factor(month,         levels = levels(train$month)),
    start_station = factor(start_station, levels = levels(train$start_station)),
    dotw_simple   = factor(dotw_simple,   levels = levels(train$dotw_simple))
  )

```

---

# Build Predictive Models

We'll build 5 models with increasing complexity to see what improves predictions.

## Model 1: Baseline (Time + Weather)

```{r model1}
contrasts(train$dotw_simple) <- contr.treatment(7)

# Now run the model
model1 <- lm(
  Trip_Count ~ hour + dotw_simple + Temperature + Precipitation,
  data = train
)

summary(model1)
```


## Model 2: Add Temporal Lags

```{r model2}
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```

## Model 3: Add Demographics

```{r model3}
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc.x + Percent_Taking_Transit.y + Percent_White.y,
  data = train
)

summary(model3)
```

## Model 4: Add Station Fixed Effects

```{r model4}
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc.x + Percent_Taking_Transit.y + Percent_White.y +
    start_station,
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

## Model 5: Add Rush Hour Interaction

```{r model5}
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + month +
    Med_Inc.x + Percent_Taking_Transit.y + Percent_White.y +
    start_station +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```



---

# Model Evaluation

## Calculate Predictions and MAE

```{r calculate_mae}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualize Model Comparison

```{r compare_models}
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# Space-Time Error Analysis

## Observed vs. Predicted

```{r obs_vs_pred}
test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

## Spatial Error Patterns

```{r spatial_errors}
# Calculate MAE by station
station_errors <- test %>%
  group_by(start_station, start_lat.x, start_lon.y) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat.x), !is.na(start_lon.y))

# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat.x, start_lon.y) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat.x), !is.na(start_lon.y))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon.y, y = start_lat.x, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon.y, y = start_lat.x, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine with better layout
library(gridExtra)
library(grid)
grid.arrange(
  p1, p2, 
  ncol = 2,
  top = textGrob(
    "Model 2 Performance: Errors vs. Demand Patterns",
    gp = gpar(fontsize = 16, fontface = "bold")
  )
)


# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon.y, y = start_lat.x, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon.y, y = start_lat.x, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
grid.arrange(
  p1, p2,
  ncol = 2
  )
p1
```

## Temporal Error Patterns

```{r temporal_errors}
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Errors and Demographics

```{r errors_demographics}
station_errors_demo <- station_errors %>%
  mutate(start_station = as.character(start_station)) %>%
  left_join(
    station_attributes %>%
      mutate(start_station = as.character(start_station)) %>%
      select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))


# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

grid.arrange(p1, p2, p3, ncol = 2)
```

---

# Model Improvement
Based on our error analysis, add 2-3 NEW features to improve the model, including Perfect biking weather, Rolling 7-day average demand, and Same hour last week.

```{r}
model6 <- lm(
  Trip_Count ~ hour + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + month +
    Med_Inc.x + Percent_Taking_Transit.y + Percent_White.y +
    start_station +
    perfect_weather + rolling7 + same_hour_last_week +
    rush_hour * weekend,
  data = train
)
summary(model6)
```

```{r}
test <- test %>%
  mutate(
    pred5 = predict(model5, newdata = test),
    pred6 = predict(model6, newdata = test)
  )

mae5 <- mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
mae6 <- mean(abs(test$Trip_Count - test$pred6), na.rm = TRUE)

mae5
mae6

```


# Critical Reflection 

1. **Operational implications:**
The final model provides reasonably accurate predictions for most station-hours and could support Indego’s rebalancing decisions, especially during low- and medium-demand periods. However, errors become more impactful during commute peaks, sudden weather changes, or holiday weeks, when shortages or surpluses are harder to correct. For this reason, the system is best used as a decision-support tool rather than a fully automated rebalancing solution. It performs well enough to guide planning, but real-time monitoring and human judgment would still be necessary.

2. **Equity considerations:**
Prediction errors are not evenly distributed across the city: some lower-income or more transit-dependent neighborhoods show higher error levels. If uncorrected, this could unintentionally reduce bike availability in communities that already face mobility constraints. To avoid reinforcing existing disparities, Indego should track errors by neighborhood, apply minimum service standards, or include equity-weighted adjustments when allocating bikes.

3. **Model limitations:**
The model cannot fully capture holiday effects, special events, or sudden shifts in winter weather, and it assumes that past patterns reliably predict future demand. It also does not account for supply-side constraints like station capacity or bike availability. With additional data—such as event calendars, improved weather forecasts, or rolling demand indicators—the model could better reflect real-world fluctuations and further improve prediction accuracy.

---
# Brief Report
For this assignment, we analyzed Q4 2024, a season with colder weather and major holidays, providing a contrast to Q1 2025 and offering insight into how seasonal factors affect bike share demand. Model comparison showed that performance improved as temporal lags, demographics, and station fixed effects were added, though Q4 predictions were more sensitive to temperature and precipitation.

Error analysis revealed that stations in high-activity areas exhibited higher variability, while peak commuting hours and holiday weeks were the most difficult to predict. Some lower-income or transit-dependent neighborhoods also showed slightly higher errors, indicating potential equity concerns if predictions directly guide re-balancing.

To improve accuracy, we added three features: a perfect-weather indicator, a rolling 7-day average, and a same-hour-last-week lag. These captured nonlinear weather effects and stable weekly routines, reducing MAE in the enhanced model, showing its advantage for decision support.

